---
title: "Chapter8 Summary: PART 3 - Questions 5,10"
output:
  html_document:
    df_print: paged
---

## Setup

```{r setup the directory}
#setwd("/Users/anjali/Documents/university/Econ_5305/code_directory/Ch8_submission_full/")
rm(list = ls())
```

```{r imports}
library(openxlsx)
library(readxl)
library(ggplot2)
library(dynlm)
library(forecast)
library(urca)
library(lubridate)
library(TSstudio)
```

```{r storage }
# create a vector to store reference to all the models
model_names <- c()
model_aic <- c()
model_bic <- c()
model_cor <- c()
model_stationary <- c()
model_invertible <-c()
```

```{r function to store stuff}
store.function <- function(name = "None",this_model,DATA,stationary = TRUE,invertible = TRUE){
  model_names <<- append(model_names,name)
  model_aic <<- append(model_aic,AIC(this_model))
  model_bic <<- append(model_bic,BIC(this_model))
  this_cor <<- cor(fitted(this_model), DATA, use="pairwise.complete.obs")^2
  model_cor<<-append(model_cor,this_cor)
  model_stationary <<- append(model_stationary,stationary)
  model_invertible <<- append(model_invertible,invertible)  
}
```

# Question 5
### 5.1 Read Data
```{r read data on GDP}
gdp_data <- read_excel("q5_gdp.xlsx")
#View(gdp_data)
GDP<-ts(gdp_data$GDP, frequency = 4, start = c(1947,4,1), end = c(2023,1,1)) 
```

### 5.2 EDA and unit root test
```{r EDA on GDP}
cat("GDP - Quarterly") 
summary(GDP)
cat("Standard deviation: ", sd(GDP))
ts.plot(GDP, main = "GDP - Quarterly", ylab="percentage")
ur.df(GDP,type="drift",lags=0)
```
**Unit Root Test: ** There is no unit root as the DF test statistic is very low. -13. It doesn't look like the series any trend.

```{r ACF and PACF anlaysis for car sales}
par(mfrow=c(1,2) , mar = c(3, 2, 3, 2))
acf(GDP,lag=40,main = "ACF: GDP")
pacf(GDP,lag=40, main = "PACF: GDP")
```
**Interpretations: ** The ACF and PACF both go up and down. This is indicative of seasonality. The PACF has spikes at lag 1 and 2. The ACF has spikes at lag = 1,2 ... 9,10,11 ... 16,17,18,19. This is indicative of ARMA models and seasonal ARMA models. We can try ARMA(1,1), ARMA(2,2), ARMA(2,4), ARMA(4,2) and auto.arima()

### Search Models

```{r Train Test Split }
# Split the data into training and testing sets.
# Keep out last 3 months
split_gdp <- ts_split(ts.obj = GDP, sample.out = 4)
train_gdp <- split_gdp$train
test_gdp <- split_gdp$test
```

```{r Car ARMA models}
ar1_ma1 <- arima(train_gdp, order=c(1, 0, 1))
ar1_ma1
ar2_ma2 <- arima(train_gdp, order=c(2, 0, 2))
ar2_ma2
ar2_ma4 <- arima(train_gdp, order=c(2, 0, 4))
ar2_ma4
ar4_ma2_d1 <- arima(train_gdp, order=c(4, 1, 2))
ar4_ma2_d1
auto_arima <- auto.arima(train_gdp, seasonal = TRUE)
auto_arima

plot(ar1_ma1, main ="ARMA (1,1): Inverse Roots")
plot(ar2_ma2, main ="ARMA(2,2): Inverse Roots")
plot(ar2_ma4, main ="ARMA(2,4): Inverse Roots")
plot(ar4_ma2_d1, main ="ARMA(4,2): Inverse Roots")
plot(auto_arima, main ="auto_arima (1,1,3)(2,0,0)[4]")

store.function("GDP ARMA(1,1)",ar1_ma1,train_gdp,TRUE,TRUE)
store.function("GDP ARMA(2,2) 1",ar2_ma2,train_gdp,TRUE,TRUE)
store.function("GDP ARMA(2,4)",ar2_ma4,train_gdp,TRUE,TRUE)
store.function("GDP ARMA(4,2) Diff 1",ar4_ma2_d1,train_gdp,TRUE,TRUE)
store.function("GDP Auto (1,1,3)(2,0,0)[4]",auto_arima,train_gdp,TRUE,TRUE)
```

```{r check all models}
all_models <- data.frame(model_names, model_aic, model_bic, model_cor, model_stationary, model_invertible)
sorted_models <- all_models[order(all_models$model_aic,decreasing = FALSE), ]
sorted_models
```

**Model Selected: ** GDP ARMA(4,2) Diff 1 and GDP Auto (1,1,3)(2,0,0)[4]

```{r GDP Forecast}
 # 3 month forecast using ARMA(4,1,2)
forecast1 <- forecast(ar4_ma2_d1, h = 4)
forecast1
plot(forecast1, main = "GDP - Forecast (ARMA (4,1,2))", include = 50)
lines(fitted(ar4_ma2_d1),col="red")
lines(GDP ) #plot the original series
legend("bottomright", legend=c("Actual", "Fitted","Forecast"), lty=c(2,2,1), col=c( "black","red", "blue"))

# 3 month forecast using Auto Arima (1,1,3)(2,0,0)[4]
forecast2 <- forecast(ar4_ma2_d1, h = 4)
forecast2
plot(forecast2, main = "GDP - Forecast (Auto (1,1,3)(2,0,0)[4])", include = 50)
lines(fitted(auto_arima),col="red")
lines(GDP) #plot the original series
legend("bottomright", legend=c("Actual", "Fitted","Forecast"), lty=c(2,2,1), col=c( "black","red", "blue"))

```

