---
title: "Chapter8 Summary: PART 3 - Questions 5,10"
output:
  html_document:
    df_print: paged
---

## Setup

```{r setup the directory}
setwd("/Users/anjali/Documents/university/Econ_5305/code_directory/Ch8_submission_full/")
rm(list = ls())
```

```{r imports}
library(openxlsx)
library(readxl)
library(ggplot2)
library(dynlm)
library(forecast)
library(urca)
library(lubridate)
library(TSstudio)
```

```{r storage }
# create a vector to store reference to all the models
model_names <- c()
model_aic <- c()
model_bic <- c()
model_cor <- c()
model_stationary <- c()
model_invertible <-c()
```

```{r function to store stuff}
store.function <- function(name = "None",this_model,DATA,stationary = TRUE,invertible = TRUE){
  model_names <<- append(model_names,name)
  model_aic <<- append(model_aic,AIC(this_model))
  model_bic <<- append(model_bic,BIC(this_model))
  this_cor <<- cor(fitted(this_model), DATA, use="pairwise.complete.obs")^2
  model_cor<<-append(model_cor,this_cor)
  model_stationary <<- append(model_stationary,stationary)
  model_invertible <<- append(model_invertible,invertible)  
}
```

# Question 10
### 10.1 Read Data
```{r read data on car sales}
data_car <- read_excel("q10_carsales.xlsx")
#View(data_car)
SALES<-ts(data_car$CAR_SALES, frequency = 12, start = c(1976,1,1), end = c(2023,3,1)) 
```

### 10.2 EDA and unit root test
```{r EDA on Car Sales}
cat("Car Sales - Monthly") 
summary(SALES)
cat("Standard deviation: ", sd(SALES))
ts.plot(SALES, main = "Time Series of Car Sales, Monthly (seasonally adjusted)", ylab="millions of unit")
ur.df(SALES,type="drift",lags=0)
```
**Unit Root Test: ** There is no unit root as the DF test statistic is very low. -5.5.

```{r ACF and PACF anlaysis for car sales}
dSALES <- diff(SALES)
dlogSALES <-diff(log(SALES))
par(mfrow=c(1,2) , mar = c(3, 2, 3, 2))
acf(SALES,lag=40,main = "ACF: Car Sales")
pacf(SALES,lag=40, main = "PACF: Car Sales")
# acf(dSALES,lag=40)
# pacf(dSALES,lag=40)
# acf(dlogSALES,lag=40)
# pacf(dlogSALES,lag=40)
```
**Interpretations: ** The ACF is smoothly decreasing while the PACF has spikes at lag = 1,2,3. AR(1) and AR(3) model is indicated.Also try the model ARMA(1,4) and ARMA(3,4) as done in the discussion.

### Search Models

```{r Train Test Split }
# Split the data into training and testing sets.
# Keep out last 3 months
split_sales <- ts_split(ts.obj = SALES, sample.out = 3)
train_sales <- split_sales$train
test_sales <- split_sales$test
```

```{r Car auto arima model}
auto_arima <- auto.arima(train_sales, seasonal = TRUE)
auto_arima
plot(auto_arima)
store.function("Auto arima ARIMA(0,1,2)(0,0,2)[12] ",auto_arima,train_sales,TRUE,TRUE)
```


```{r Car AR models}
car_ar1 <- arima(train_sales, order=c(1, 0, 0))
car_ar1
car_ar1_d1 <- arima(train_sales, order=c(1, 1, 0))
car_ar1_d1
car_ar3 <- arima(train_sales, order=c(3, 0, 0))
car_ar3
car_ar3_d1 <- arima(train_sales, order=c(3, 1, 0))
car_ar3_d1

par(mfrow=c(2,2) , mar = c(3, 2, 3, 2))
plot(car_ar1, main ="AR1: Inverse Roots")
plot(car_ar1_d1, main ="AR1 (Diff 1): Inverse Roots")
plot(car_ar3, main ="AR3: Inverse Roots")
plot(car_ar3_d1, main ="AR3 (Diff 1): Inverse Roots")

store.function("Car Sales AR1",car_ar1,train_sales,TRUE,TRUE)
store.function("Car Sales AR1 Diff 1",car_ar1_d1,train_sales,TRUE,TRUE)
store.function("Car Sales AR3",car_ar3,train_sales,FALSE,TRUE)
store.function("Car Sales AR3 Diff 3",car_ar3_d1,train_sales,TRUE,TRUE)
```
```{r Car ARMA models}
car_ar1_ma4 <- arima(train_sales, order=c(1, 0, 4))
car_ar1_ma4
car_ar1_ma4_d1 <- arima(train_sales, order=c(1, 1, 4))
car_ar1_ma4_d1

car_ar3_ma4 <- arima(train_sales, order=c(3, 0, 4))
car_ar3_ma4
car_ar3_ma4_d1 <- arima(train_sales, order=c(3, 1, 4))
car_ar3_ma4_d1

plot(car_ar1_ma4, main ="ARMA(1,4): Inverse Roots")
plot(car_ar1_ma4_d1, main ="ARMA(1,4) Diff 1: Inverse Roots")
plot(car_ar3_ma4, main ="ARMA(3,4): Inverse Roots")
plot(car_ar3_ma4_d1, main ="ARMA(3,4) Diff 1: Inverse Roots")

store.function("Car Sales ARMA(1,4)",car_ar1_ma4,train_sales,FALSE,TRUE)
store.function("Car Sales ARMA(1,4) Diff 1",car_ar1_ma4_d1,train_sales,TRUE,TRUE)
store.function("Car Sales ARMA(3,4)",car_ar3_ma4,train_sales,FALSE,FALSE)
store.function("Car Sales ARMA(3,4) Diff 1",car_ar3_ma4_d1,train_sales,TRUE,TRUE)
```

```{r check all models}
all_models <- data.frame(model_names, model_aic, model_bic, model_cor, model_stationary, model_invertible)
sorted_models <- all_models[order(all_models$model_aic,decreasing = FALSE), ]
sorted_models
```

**Model Selected: ** 1.Auto arima ARIMA(0,1,2)(0,0,2)[12] 2. Car Sales ARMA(1,4) Diff 1 -> ARIMA(1,1,4)
```{r Car Sales Forecast}
# Auto arima ARIMA(0,1,2)(0,0,2)[12]
forecast1 <- forecast(auto_arima, h = 3)
forecast1
plot(forecast1, main = "Car Sales Units - Forecast (ARIMA(0,1,2)(0,0,2)[12])", include = 50)
lines(fitted(auto_arima),col="red")
lines(SALES) #plot the original series
legend("bottomright", legend=c("Actual", "Fitted","Forecast"), lty=c(1,1,1), col=c( "black","red", "blue"))

#3 month forecast using ARMA(1,1,4)
forecast2 <- forecast(car_ar1_ma4_d1, h = 3)
forecast2
plot(forecast2, main = "Car Sales Units - Forecast (ARMA (1,1,4))", include = 50)
lines(fitted(car_ar1_ma4_d1),col="red")
lines(SALES) #plot the original series
legend("bottomright", legend=c("Actual", "Fitted","Forecast"), lty=c(1,1,1), col=c( "black","red", "blue"))
```

The auto arima picks up the seasonality very well and creates a very good model that can be used for forecasting. 